Highly Available Transactions: Virtues and Limitations

### 2013-03-25 writeup/evaluation.tex

As we have discussed, many existing database transaction and
concurrency control algorithms are not designed for high
availability. These algorithms often presume a single-server
deployment or the requirement for serializability. As a consequence,
traditional transaction processing systems are not well-optimized for
a HAT context.

@@

\item \textbf{Optimistic concurrency control} allows reads from any
  server but transactions must be checked by a scheduler before
  commit. With one scheduler per database, this requires one RTT for
  concurrency control. If schedulers are sharded by database
  partition, a multi-partition transaction may require contacting
  several different schedulers and, possibly, an atomic commit
  protocol, which becomes substantially more expensive.

\item \textbf{Deterministic transaction scheduling} is effecitvely the
  inverse of optimistic concurrency control: a scheduler first orders
  the transactions, then sends them to database replicas for in-order
  execution. These techniques, pioneered in the context of totally
  ordered atomic broadcast systems, have recently seen increased
  attention and effectively require communication overhead similar to
  optimistic concurrency control.

@@

There are several mechanisms that do not provide serializability (and
often provide HAT semantics) but which are not
available. Non-serializability can but does not necessarily imply high
availability and low latency:

\begin{itemize}
\item \textbf{Lock-based protocols} such as those used to provide weak
  isolation in Gray's original paper require at least one RTT to the a
  lock server (or, alternatively, a majority quorum). In the presence
  of failure, distributed locking can be challenging.
\item The \textbf{MDCC} commit protocol uses a Paxos variant and escrow
  transactions to provide Repeatable Read isolation with 1-2 RTTs. (TODO: CHECK)
\item \textbf{Parallel Snapshot Isolation} as developed by the Walter
  system provides (unavailable) snapshot isolation variant and, even
  when writes commute, requires 1 RTT to enforce read
  recency. However, PSI without recency requirements could be made
  highly available.
\item \textbf{Eiger} provides read-only and write-only transactions
  with item cut isolation and causal consistency. It relies on sticky
  high availability and groups within a datacenter with linearizable
  clusters. In the presence of server failure during write
  transactions, reads and writes may become unavailable. \textbf{COPS}
  provides similar semantics for read-only transactions but does not
  suffer from unavailability as it does not have a coordinator.
\item \textbf{Chan and Gray read-only transactions} use a similar
  algorithm to Eiger to provide read-only transactions with item cut
  isolation and causal consistency but coordinate on reads instead of
  writes.
\item \textbf{Bolt-on Causal Consistency} and \textbf{Swift} can
  provide read-only and write-only transactions with item cut
  isolation and causal consistency with sticky high availability and
  both rely on client-side caching.
\item \textbf{Bayou} provides several session guarantees but, without
  client-side caching, actually faces unavailability in the presence
  of failure. (TODO: CHECK)
\item \textbf{Brantner's S3 Database} uses caching to provide several
  guarantees such as XX and YY but is unavailable during failures
  (TODO: CHECK).
\end{itemize}

@@

Thus far, we have largely ignored the question of which applications
actually require HAT semantics versus traditional ACID guarantees. As
we summarized in Section~\ref{needed}, the main cost of providing HAT
updates is the inability to prevent Lost Update, Write Skew, and
provide recency bounds. In this Section, we attempt to understand when
these guarantees matter both abstractly and by the use of a
representative customer order transaction application patterned off of
TPC-C.

@@

Absent a necessary and sufficient result describing exactly which
application logics can be achieved with high availability and to make
these requirements more concrete, we consider an example application:
the TPC-C benchmark. Our goal is \textit{not} to provide a performance
analysis, as any HAT configuration will be non-compliant. Rather, we
examine the five queries contained in the benchmarks---ostensibly
representative of a standard transaction processing application---to
understand which are achievable in a HAT context.

TPC-C consists of five transactions, capturing operation of a
real-world warehouse, including sales, payments, and deliveries. We
analyze each transaction and its interactions below.

\begin{itemize}
\item \textbf{New-Order} places an order for a variable quantity of
  data items, updating warehouse stock as needed. The New-Order
  transaction selects a sales district, assigns the new order an order
  ID number, adjusts the remaining warehouse stock, and writes a
  placeholder entry for the pending order. There are two mutations in
  this transaction which we must be careful of.

  First, assigning an order ID number is challenging. We can easily
  assign a \textit{unique} order ID, by, say, concatenating the user
  ID with a timestamp and server ID. However, the TPC-C specification
  requires that order numbers be sequentially assigned, which we
  cannot guarantee in HATs due to Lost Update. Accordingly, while HATs
  cannot provide compliant TPC-C execution due to a non-monotonic
  counter increase, they can allow placement of new orders, which is a
  commutative and monotonic operation

  Second, decrementing inventory counts is a potentially risky
  operation as, without coordination, inventory counts may become
  negative. However, TPC-C automatically ``refills'' inventory counts
  (by 91 items) when there are 10 or fewer items
  remaining. Accordingly, even in the presence of concurrent new
  orders to the same item, the stock will never fall below
  zero. However, if the system propagates ``refill'' operations
  between replicas, then an $R$ replica system might end up with
  $91(R-1)$ more stock than a single-site implementation.

>>>>

Many existing database transaction and concurrency control algorithms
are not designed for high availability. These algorithms often presume
a single-server deployment or the requirement for serializability. As
a consequence, traditional transaction processing systems are not
well-optimized for a HAT context. In this section, we briefly discuss
design decisions and algorithmic details that preclude high
availability.

@@

\item \textbf{Optimistic concurrency control} allows reads from any
  server but transactions must be checked by a scheduler before
  commit. With one scheduler per database, this requires one RTT
  specifically for concurrency control. If schedulers are sharded by
  database partition, a multi-partition transaction may require
  contacting several different schedulers and, possibly, an atomic
  commit protocol, which becomes substantially more expensive.

\item \textbf{Deterministic transaction scheduling} is ffectively the
  inverse of optimistic concurrency control: a scheduler first orders
  the transactions, then sends them to database replicas for in-order
  execution. These techniques, pioneered in the context of totally
  ordered atomic broadcast systems, have recently seen increased
  attention and effectively require communication overhead similar to
  optimistic concurrency control.

@@

Perhaps unsurprisingly, the reliance on a glocal total order
necessitates a minimum of one round-trip to a designated master or
coordination service for each of these classic algorithms. This is
simply a cost of providing serializability, and the cost of the
(minimum one) round trip will be determined by the deployment
environment, as we saw in Section~\ref{sec:motivation}.

There also are several mechanisms that do not provide serializability
but provide transactional semantics with varying availability
properties:

\begin{itemize}
\item \textbf{Lock-based protocols} such as those used to provide weak
  isolation in Gray's original paper require at least one RTT to the a
  lock server (or, alternatively, a majority quorum). In the presence
  of failure, distributed locking is challenging.
\item The \textbf{MDCC} commit protocol uses a Paxos variant and escrow
  transactions to provide Repeatable Read isolation with 1-2 RTTs. (TODO: CHECK)
\item \textbf{Parallel Snapshot Isolation} as developed by the Walter
  system provides (unavailable) snapshot isolation variant and, even
  when writes commute, requires 1 RTT to enforce read
  recency. However, PSI without recency requirements could be made
  highly available.
\item \textbf{Eiger} provides read-only and write-only transactions
  with item cut isolation and causal consistency. It relies on sticky
  high availability and groups within a datacenter with linearizable
  clusters. In the presence of server failure during write
  transactions, reads and writes may become unavailable. \textbf{COPS}
  provides similar semantics for read-only transactions but does not
  suffer from unavailability as it does not have a coordinator.
\item \textbf{Chan and Gray's Read-only transactions} use a similar
  algorithm to Eiger to provide read-only transactions with item cut
  isolation and causal consistency but coordinate on reads instead of
  writes.
\item \textbf{Bolt-on Causal Consistency} and \textbf{Swift} can
  provide read-only and write-only transactions with item cut
  isolation and causal consistency with sticky high availability and
  both rely on client-side caching.
\item \textbf{Bayou} provides several session guarantees but, without
  client-side caching, actually faces unavailability in the presence
  of failure. (TODO: CHECK)
\item \textbf{Brantner's S3 Database} uses caching to provide several
  guarantees such as XX and YY but is unavailable during failures
  (TODO: CHECK).
\end{itemize}

@@

Thus far, we have largely ignored the question of which applications
require semantics which are unavailable to HATs. As we showed in
Section~\ref{sec:hats}, the main cost of choosing HATs comes in the
inability to prevent Lost Update, Write Skew, and provide recency
bounds. In this Section, we attempt to understand when these
guarantees matter both abstractly and by the use of a representative
customer order transaction application patterned off of TPC-C.

@@

As an example of an actual system rearchitecture,
less than one year after its announcement, Yahoo!'s PNUTS developers
explicitly added a support for weaker, highly available operation,
explaining that ``strict adherence [to strong consistency] leads to
difficult situations under network partitioning or server
failures...in many circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

Absent a necessary and sufficient result describing exactly which
application logics can be achieved with high availability and to make
these requirements more concrete, we consider an example application:
the TPC-C benchmark. Our goal is \textit{not} to provide a performance
analysis, as any HAT configuration will be non-compliant. Rather, we
examine the five queries contained in the benchmarks---ostensibly
representative of a standard transaction processing application---to
understand which are achievable in a HAT context. TPC-C consists of
five transactions, capturing operation of a real-world warehouse,
including sales, payments, and deliveries. We analyze each transaction
and its interactions below.

\begin{itemize}
\item \textbf{New-Order} places an order for a variable quantity of
  data items, updating warehouse stock as needed. The New-Order
  transaction selects a sales district, assigns the new order an order
  ID number, adjusts the remaining warehouse stock, and writes a
  placeholder entry for the pending order. There are two mutations we
  must be consider in detail.

  First, assigning an order ID number is challenging. We can easily
  assign a \textit{unique} order ID, by, say, concatenating the
  customer ID with a timestamp and server ID. However, the TPC-C
  specification requires that order numbers are assigned sequentially,
  which we cannot guarantee in HATs due to Lost Update. Accordingly,
  HATs can allow placement of new orders with unique identifiers but
  cannot provide compliant TPC-C execution due to the non-monotonic
  counter increase.

  Second, decrementing inventory counts is a potentially risky
  operation as, without coordination, inventory counts may become
  negative. However, TPC-C automatically ``restocks'' inventory counts
  (by 91 items) when there are 10 or fewer items
  remaining. Accordingly, even in the presence of concurrent new
  orders to the same item, the stock will never fall below
  zero. However, an $R$ replica HAT system might end up with $91(R-1)$
  more stock than a single-site implementation.
